Inference methods comparison for the simplified Naomi model using Pareto-smoothed importance sampling

### 2023-05-11 src/naomi-simple_psis/psis.Rmd
First try at producing PSIS k hat statistic #41

```{r}
tmbstan_samples <- rstan::extract(tmbstan$mcmc$stanfit)

data.frame(
  y = tmbstan_samples$lp__,
  x = 1:length(tmbstan_samples$lp__)
) %>%
  ggplot(aes(x = x, y = y)) +
    geom_point(alpha = 0.4) +
    labs(y = "Log-likelihood", x = "Draw number") +
    theme_minimal() 
```

Need to write a function which reformats `TMB` or `tmbstan` samples to be evaluated using `objfull$fn`:

```{r}
objfull$par

#' TODO
```

>>>>

```{r}
tmbstan_extract <- rstan::extract(tmbstan$mcmc$stanfit)

data.frame(
  y = tmbstan_extract$lp__,
  x = 1:length(tmbstan_extract$lp__)
) %>%
  ggplot(aes(x = x, y = y)) +
    geom_point(alpha = 0.4) +
    labs(y = "Log-likelihood", x = "Draw number") +
    theme_minimal()
```

We would like to produce evalautions of the log-likelihood for the `TMB` and `aghq` samples as well.
To do this, we first create a function which reformats the samples to be evaluated using `objfull$fn`:

```{r}
par_samp_matrix <- function(sample, elements = unique(names(objfull$par))) {
  # Keep and reorder to be the same as elements
  x <- sample[elements]
  # Bind rows together
  do.call(rbind, x)
}

tmb_samples <- par_samp_matrix(tmb$fit$sample)
aghq_samples <- par_samp_matrix(aghq$quad$sample)
tmbstan_samples <- par_samp_matrix(tmbstan$mcmc$sample)
```

Now we can calculate the log-likelihood by applying this function across the rows of the sample matrices:

```{r}
tmb_ll <- apply(tmb_samples, 2, FUN = function(x) -1 * objfull$fn(x))
aghq_ll <- apply(aghq_samples, 2, FUN = function(x) -1 * objfull$fn(x))
tmbstan_ll <- apply(tmbstan_samples, 2, FUN = function(x) -1 * objfull$fn(x))

data.frame(
  y = c(tmb_ll, aghq_ll, tmbstan_ll),
  x = c(1:length(tmb_ll), 1:length(aghq_ll), 1:length(tmbstan_ll)),
  method = c(rep("TMB", times = length(tmb_ll)), rep("aghq", times = length(aghq_ll)), rep("tmbstan", times = length(tmbstan_ll)))
) %>%
  ggplot(aes(x = x, y = y, col = method)) +
    geom_point(alpha = 0.4) +
    facet_grid(~ method, scales = "free_x") +
    scale_colour_manual(values = c("#56B4E9", "#009E73", "#E69F00")) +
    labs(y = "Log-likelihood", x = "Draw number", col = "Method") +
    theme_minimal()
```

To check that this is working as expected, we can compare the answers produced this way to those obtained directly from `lp__` in Stan:

```{r}
data.frame(lp__ = sort(tmbstan_extract$lp__), objfull = sort(tmbstan_ll)) %>%
  ggplot(aes(x = lp__, y = objfull)) +
    geom_point(alpha = 0.4) +
    labs(title = "Appears these points lie on a straight line") +
    theme_minimal()
```

It's interesting to note the differing levels of the log-likelihoods between the three sampling methods.
Is there a way that this could be interpreted?
Stan samples from the typical set, whereas `aghq` and `TMB` are sampling from closer to the mode?

```{r}
-1 * mean(tmb_ll)
-1 * mean(aghq_ll)
-1 * mean(tmbstan_ll)

(mean(tmb_ll) - mean(tmbstan_ll)) / 24
```

To run PSIS we need to get the log-likelihoods under the proposal, as well as the log-likelihoods under the target.
The proposal for `TMB` is a multivariate Gaussian, but only for the random effects:

```{r}
r <- tmb$fit$obj$env$random # Indices of the random effects
par_r <- tmb$fit$par.full[r] # Mode of the random effects

# Hessian of the random effects (this has crashed on me before, so being safe)
safe <- parallel::mcparallel({tmb$fit$obj$env$spHess(tmb$fit$par.full, random = TRUE)})
hess_r <- parallel::mccollect(safe, wait = TRUE, timeout = 0, intermediate = FALSE)
```

For `aghq` it's a mixture of Gaussians:

```{r}
aghq_modes <- aghq$quad$modesandhessians$mode
aghq_hessians <- aghq$quad$modesandhessians$H
aghq_lambda <- exp(aghq$quad$normalized_posterior$nodesandweights$logpost_normalized) * aghq$quad$normalized_posterior$nodesandweights$weights
```

`aghq_lambda` contains the multinomial weights, which we verify are in fact weights here:

```{r}
stopifnot(sum(aghq_lambda) - 1 < 10E-9)
stopifnot(all(aghq_lambda > 0))
```

Now to construct the log-likelihood, we will use the log-sum-exp trick:

```{r}
aghq_chol <- lapply(aghq_hessians, Matrix::Cholesky)
aghq_latent_samples <- par_samp_matrix(aghq$quad$sample, elements = unique(names(aghq_modes[[1]])))

start <- Sys.time()
mvn_ll <- mapply(sparseMVN::dmvn.sparse, mu = aghq_modes, CH = aghq_chol, MoreArgs = list(x = t(aghq_latent_samples), prec = TRUE, log = TRUE))
end <- Sys.time()

#' Takes around 2 minutes
end - start

aghq_q_ll <- apply(mvn_ll, 1, FUN = function(row) matrixStats::logSumExp(log(aghq_weights) + row))
plot(aghq_q_ll)
```

Now run the PSIS with `aghq`.
The importance ratios are $\log(p(\theta, y)) - \log(q(\theta))$:

```{r}
r <- - aghq_ll - aghq_q_ll
plot(r)

aghq_psis <- loo::psis(log_ratios = lw, r_eff = 1)
aghq_psis$diagnostics
```
